---
title: "Categorical inputs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{categorical_inputs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(123)
```

In this example, we will illustrate several ways of dealing with categorical variables when using grf.

One of the approaches below relies on `grf`'s sister package [`sufrep`](https://github.com/grf-labs/sufrep). Let's install and load it first.

```{r setup}
library(grf)
install.packages("https://github.com/grf-labs/sufrep/blob/master/sufrep_0.1.0.tar.gz?raw=true", repos = NULL, type = "source")
library(sufrep)
```

Let's pretend we would like to estimate mileage per gallon (`mpg`) from number of cylinders (`cyl`),
quarter-mile time (`qsec`), and car brand name (`brand`, created below).

```{r dataset}
# Create a categorical column with brand name
df <- within(mtcars, {
  # E.g. 'Mazda RX4' --> 'Mazda'
  brand <- factor(sapply(rownames(mtcars), function(x) strsplit(x, " ")[[1]][1]))
})

x <- c("cyl", "qsec") # Continuous variables
g <- c("brand")       # Categorical variable

head(df[c(x, g)])
```

This code would raise an error, because data is not numerical.

```{r problem}
# rf <- regression_forest(X=df[c(x, g)], Y=df$mpg)
```

We can consider three approaches here.

+ Simply assign integers to each category (convert 'AMC' to 1, 'Cadillac' to 2, etc.)
+ One-hot encode the categories (as many binary columns as there are categories)
+ Use a _sufficient representation_ of the category. Here we will use the `means` method from the [`sufrep`](https://github.com/grf-labs/sufrep) package.

The last method involves substituting the `brand` column by averages of the continuous columns `cyl` and `qsec`, grouped by category. If you are curious about why that works, or would like to know more about sufficient representations, please check out our `sufrep` paper ([ArXiv](https://arxiv.org/abs/1908.09874v1)).

```{r solutions}
# Solution 1: Transform variable into numbers
X1 <- within(df[c(x, g)], brand <- as.numeric(brand))
rf1 <- regression_forest(X1, df$mpg)


# Solution 2: One-hot encoding
X2 <- model.matrix(~ 0 + ., df[c(x, g)])
rf2 <- regression_forest(X2, df$mpg)


# Solution 3: 'Means' encoding using the 'sufrep' package
encoder <- make_encoder(df[x], df$brand, method="means")
X3 <- encoder(df[x], df$brand)
rf3 <- regression_forest(X3, df$mpg)
```

Different approaches can yield different forest performance.

```{r mses}
mse1 <- mean(rf1$debiased.error)
mse2 <- mean(rf2$debiased.error)
mse3 <- mean(rf3$debiased.error)

print("MSE when representing categorical variables as...")
print(paste0("Integers: ", mse1))
print(paste0("One-hot vectors: ", mse2))
print(paste0("'Means' encoding [sufrep]: ", mse3))
```
