---
title: "Evaluating a causal forest fit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(123)
```

```{r setup}
library(grf)
```

Two common diagnostics to evaluate if the identifying assumptions behind grf hold is a propensity score histogram and covariance balance plot.

```{r}
n <- 2000
p <- 10
X <- matrix(rnorm(n * p), n, p)

W <- rbinom(n, 1, 0.4 + 0.2 * (X[, 1] > 0))
Y <- pmax(X[, 1], 0) * W + X[, 2] + pmin(X[, 3], 0) + rnorm(n)
cf <- causal_forest(X, Y, W)
```

The overlap assumptions requires a positive probability of treatment for each $X_i$. We should not be able to deterministically decide the treatment status of an individual based on its covariates, meaning none of the estimated propensity scores should be close to one or zero. One can check this with a histogram:

```{r}
hist(e.hat <- cf$W.hat)
```

One can also check that the covariates are balanced across the control and treated group by plotting the histogram $X_i | W_i = k$, overlaid here for each feature:

```{r, fig.height = 25}
par(mar = c(5, 5, 1, 1), oma = c(0, 0, 2, 0), mfrow = c(p/2, 2))
plots <- lapply(1:p, function (i) {
  control.hist <- hist(X[W == 0, i], plot = FALSE)
  treated.hist <- hist(X[W == 1, i], plot = FALSE)

  plot(control.hist, col= rgb(1, 0, 0, 0.5), main = paste("variable", i), xlab = "")
  plot(treated.hist, col= rgb(0, 0, 1, 0.5), add = TRUE)
})
title("Covariate histograms, control/treated (red/blue)", outer = TRUE)
```
