---
title: "Introduction to grf"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{grf}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(123)
```

```{r setup}
library(grf)
```

The following example demonstrates how to use GRF for heterogeneous treatment effect estimation. For examples of how to use types of forest, as for quantile regression and causal effect estimation using instrumental variables, please consult the R documentation on the relevant forest methods (quantile_forest, instrumental_forest, etc.).

Generate data and train a causal forest
```{r}
n <- 2000
p <- 10
X <- matrix(rnorm(n * p), n, p)
X.test <- matrix(0, 101, p)
X.test[, 1] <- seq(-2, 2, length.out = 101)

W <- rbinom(n, 1, 0.4 + 0.2 * (X[, 1] > 0))
Y <- pmax(X[, 1], 0) * W + X[, 2] + pmin(X[, 3], 0) + rnorm(n)
tau.forest <- causal_forest(X, Y, W)
tau.forest
```

Estimate treatment effects for the training data using out-of-bag prediction.
```{r}
tau.hat.oob <- predict(tau.forest)
hist(tau.hat.oob$predictions)
```

Estimate treatment effects for the test sample.
```{r}
tau.hat <- predict(tau.forest, X.test)
plot(X.test[, 1], tau.hat$predictions, ylim = range(tau.hat$predictions, 0, 2), xlab = "x", ylab = "tau", type = "l")
lines(X.test[, 1], pmax(0, X.test[, 1]), col = 2, lty = 2)
```

Estimate the conditional average treatment effect on the full sample (CATE).
```{r}
average_treatment_effect(tau.forest, target.sample = "all")
```

Estimate the conditional average treatment effect on the treated sample (CATT).
Here, we don't expect much difference between the CATE and the CATT, since
treatment assignment was randomized.
```{r}
average_treatment_effect(tau.forest, target.sample = "treated")
```

Add confidence intervals for heterogeneous treatment effects; growing more trees is now recommended.
```{r}
tau.forest <- causal_forest(X, Y, W, num.trees = 4000)
tau.hat <- predict(tau.forest, X.test, estimate.variance = TRUE)
sigma.hat <- sqrt(tau.hat$variance.estimates)
plot(X.test[, 1], tau.hat$predictions, ylim = range(tau.hat$predictions + 1.96 * sigma.hat, tau.hat$predictions - 1.96 * sigma.hat, 0, 2), xlab = "x", ylab = "tau", type = "l")
lines(X.test[, 1], tau.hat$predictions + 1.96 * sigma.hat, col = 1, lty = 2)
lines(X.test[, 1], tau.hat$predictions - 1.96 * sigma.hat, col = 1, lty = 2)
lines(X.test[, 1], pmax(0, X.test[, 1]), col = 2, lty = 1)
```

In some examples, pre-fitting models for Y and W separately may
be helpful (e.g., if different models use different covariates).
In some applications, one may even want to get Y.hat and W.hat
using a completely different method (e.g., boosting).

```{r}
# Generate new data.
n <- 4000
p <- 20
X <- matrix(rnorm(n * p), n, p)
TAU <- 1 / (1 + exp(-X[, 3]))
W <- rbinom(n, 1, 1 / (1 + exp(-X[, 1] - X[, 2])))
Y <- pmax(X[, 2] + X[, 3], 0) + rowMeans(X[, 4:6]) / 2 + W * TAU + rnorm(n)

forest.W <- regression_forest(X, W, tune.parameters = TRUE)
W.hat <- predict(forest.W)$predictions

forest.Y <- regression_forest(X, Y, tune.parameters = TRUE)
Y.hat <- predict(forest.Y)$predictions

forest.Y.varimp <- variable_importance(forest.Y)

# Note: Forests may have a hard time when trained on very few variables
# (e.g., ncol(X) = 1, 2, or 3). We recommend not being too aggressive
# in selection.
selected.vars <- which(forest.Y.varimp / mean(forest.Y.varimp) > 0.2)

tau.forest <- causal_forest(X[, selected.vars], Y, W,
                            W.hat = W.hat, Y.hat = Y.hat,
                            tune.parameters = TRUE)
```

Check whether causal forest predictions are well calibrated.
```{r}
test_calibration(tau.forest)
```
