---
title: "Visualize trees"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{visualize_tree}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(123)
```

```{r setup}
library(grf)
```

## Visualize trees in a random forest

All GRF forest objects works with the tree extractor function `get_tree` which you can use to visually inspect trees with either `print` or `plot`. The example below demonstrates this for a regression forest.

```{r}
n <- 100
p <- 5
X <- matrix(rnorm(n * p), n, p)
Y <- X[, 1] * rnorm(n)
r.forest <- regression_forest(X, Y, num.trees = 100)

# Extract the first tree from the fitted forest.
tree <- get_tree(r.forest, 1)
# Print the first tree.
print(tree)

# Plot the first tree.
plot(tree)
```

To find the leaf node a given sample falls into, you can use the `get_leaf_node` function.

```{r}
n.test <- 4
X.test <- matrix(rnorm(n.test * p), n.test, p)
print(X.test)
# Get a vector of node numbers for each sample.
get_leaf_node(tree, X.test)
# Get a list of samples per node.
get_leaf_node(tree, X.test, node.id = FALSE)
```

For a tutorial on visualizing tree-based treatment assignment rules, see the `Policy learning` vignette.
