---
title: "Visualize trees"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{visualize_tree}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(123)
```

```{r setup}
library(grf)
```

## Visualize trees in a random forest

All GRF forest objects works with the tree extractor function `get_tree` which you can use to visually inspect trees with either `print` or `plot`. The example below demonstrates this for a regression forest.

```{r}
n <- 100
p <- 5
X <- matrix(rnorm(n * p), n, p)
Y <- X[, 1] * rnorm(n)
r.forest <- regression_forest(X, Y, num.trees = 100)

# Extract the first tree from the fitted forest.
tree <- get_tree(r.forest, 1)
# Print the first tree.
print(tree)

# Plot the first tree.
plot(tree)
```

To find the leaf node a given sample falls into, you can use the `get_leaf_node` function.

```{r}
n.test <- 4
X.test <- matrix(rnorm(n.test * p), n.test, p)
print(X.test)
# Get a vector of node numbers for each sample.
get_leaf_node(tree, X.test)
# Get a list of samples per node.
get_leaf_node(tree, X.test, node.id = FALSE)
```

## Visualizing tree-based treatment assignment rules

The above approach is only a convenient way to inspect individual trees, it is not suggested as a way to evaluate or design treatment assignment rules. For this purpose we suggest the companion package [policytree](https://github.com/grf-labs/policytree) (Athey and Wager, 2017). The example below illustrates this by fitting a shallow tree on doubly robust treatment effect estimates obtained from a causal forest. The function `policy_tree` and `double_robust_scores` belong to the [policytree](https://github.com/grf-labs/policytree) package.

```{r}
library(policytree)

# Fit a causal forest.
n <- 15000
p <- 5
X <- round(matrix(rnorm(n * p), n, p), 2)
W <- rbinom(n, 1, 1 / (1 + exp(X[, 3])))
tau <- 1 / (1 + exp((X[, 1] + X[, 2]) / 2)) - 0.5
Y <- X[, 3] + W * tau + rnorm(n)
c.forest <- causal_forest(X, Y, W)

# Compute doubly robust scores.
dr.scores <- double_robust_scores(c.forest)
# Fit a depth two tree on the doubly robust scores.
tree <- policy_tree(X, dr.scores, 2)
plot(tree)

# Predict treatment assignment.
predicted <- predict(tree, X)

plot(X[, 1], X[, 2], col = predicted)
legend("topright", c("control", "treat"), col = c(1, 2), pch = 19)
abline(0, -1, lty = 2)
```

For more details please see the referenced package, and the references therein.

## References
Susan Athey and Stefan Wager. Efficient Policy Learning. 2017. [[arxiv](https://arxiv.org/abs/1702.02896)]

Sverdrup, Erik, Ayush Kanodia, Zhengyuan Zhou, Susan Athey, and Stefan Wager. policytree: Policy learning via doubly robust empirical welfare maximization over trees. _Journal of Open Source Software 5, no. 50 (2020): 2232._ [[paper](https://joss.theoj.org/papers/10.21105/joss.02232.pdf)]
